Multi-stage programming with functors and monads: eliminating abstraction overhead from generic code

### 2007-07-07 SCP2008/paper/metamonads.tex
editing of the latter sections

Reviewing the various aspects outlined in section~\ref{aspects},
the simplest parametrization is to make the domain abstract. 
\jacques{This whole section needs to be edited to take into account
the different organization of the material}.
As it
turns out, we need the following to exist in our domains: $0$, $1$,
$+$, $*$, (unary and binary) $-$, at least \emph{exact} division,
normalization, and potentially a relative size measure. The simplest
case of such domain abstraction is |param_code1|.
There, code-generators such as |plus| and |one|
were passed as arguments. We need far more than
two parameters, so we have to group them. Instead of the grouping
offered by regular records, we use OCaml \emph{structures} (i.e.,
modules)
so we can take advantage of extensibility, type abstraction and constraints,
and especially parameterized structures (\emph{functors}).
We define the type of the domain, the signature |DOMAIN|, which
different domains must satisfy:

@@

module IntegerDomain : DOMAIN = struct
    type v = int
    let kind = Domain_is_Ring
    let zero = 0
    let one = 1
    let plus x y = x + y
    let times x y = x * y
    let minus x y = x - y
    let uminus x = -x
    let div x y = x / y
    let normalizer = None
    let better_than = Some (fun x y -> abs x > abs y)
end
\end{code}

One particular domain instance is |IntegerDomain|. The notation\\
|module IntegerDomain : DOMAIN| makes the compiler verify that our
|IntegerDomain| is indeed a |DOMAIN|, that is, satisfies the required
signature. The constraint |DOMAIN| may be omitted; in that case, the
compiler will verify the type when we try to use that structure as a
|DOMAIN|. In any case, the errors such as missing ``methods'' or
methods with incorrect types will be caught statically, even
\emph{before} any code generation takes place. The variant
|domain_is_ring| of |domain_kind| encodes a semantic constraint 
that the full division
is not available. While the |DOMAIN| type may have looked daunting to
some, the implementation is quite straightforward.  Other domains such
as |float| and arbitrary precision exact rational numbers |Num.num|
are equally simple.

A more complex domain: |Zp|
\begin{code}
module ZpMake(P:sig val p:int end) = struct
    type v = int
    let kind = Domain_is_Field
    let zero = 0 and one = 1
    let plus x y = (x + y) mod P.p
    let times x y = (x * y) mod P.p
    \dots
    let normalizer = None and better_than = None
    let () = assert (is_prime P.p)
end
\end{code}

This domain is parametrized by a an integer number |p|, the
(modulus?). Thus, ZpMake is a functor. It creates a domain that is a
field, but with no defined order. Hence
|normalizer| and |better_than| are set to |None|. However, |Zp| forms
a field only when |p| is prime. Therefore, we must make this check,
see the last line of that code. The expression 
|assert (is_prime P.p)| is the initializing expression. It is executed
when the correspo0nding module is instantaited
\begin{code}
module Z19 = ZpMakeL(struct let p = 19 end)
\end{code}
if we replace |19| with |9| above, we receive an error. That is a
run-time error. It is raised however as we instantiate and
combine modules that will eventually make the generator. Thus,
although the error is reported at run-time rather than compile time as
one might have hoped, the error is raised when \emph{generating the
generator} -- well before the generation of the target code could
begin. In our code we make extensive use of these so-called preflight 
checks performed as part of module initialization. The checks seem to
offer a good compromise: they are dynamic and so do not require
complicated type systems; on the other hand, the checks are performed
quite early, when buiding code generators, and so ensure that no 
code violating the corresponding \emph{semantic} constraints 
will be generated. Although some may frown on the use of the module
initializing expressions (since that requires careful attention to the
sharing and problems of multiple instantiation of a module), these
concerns do not apply to our case: our preflight tests are all idempotent
and maintain no state.

Lifting:

We now define a lot of infrastructure.  We will be quite thorough, and
define base abstraction and lifted abstractions, This will involve a
lot of boilerplate code, which unfortunately cannot be so easily
automated in MetaOCaml -- that would require introspection.  It could
be done in camlp4, but that seems too much as well.  This 'base' could
be elided, but when we decide to make more use of Abstract
Interpretation, we'll regret it, so do it now.
\begin{code}
let lift x = .< x >.

let liftRef x = .< ref .~x >. 
let liftGet x = .< ! .~x >. 
let unitL = fun s k -> k s .< () >.

let liftPair x = (.< fst .~x >., .< snd .~x >.)

(* logic code combinators - plain and monadic *)
module Logic = struct
  let notL a        = .< not .~a >.
  let equalL a b    = .< .~a = .~ b >.
  let notequalL a b = .< .~a <> .~ b >.
  let andL a b     = .< .~a && .~b >. 
end

(* operations on code indices *)
module Idx = struct
  let zero = .< 0 >.
  let one = .< 1 >.
  let minusone = .< -1 >.
  let succ a = .< .~a + 1 >.
  let pred a = .< .~a - 1 >.
  let less a b = .< .~a < .~b >.
  let uminus a = .< - .~a >.
  let add a b = .< .~a + .~b >.

let cunit = .< () >.
let update a f = let b = f (liftGet a) in .< .~a := .~b >.
let assign a b = .< .~a := .~b >.
let apply  f x = .< .~f .~x >.
let updateM a f = ret (update a f)
let assignM a b = ret (assign a b)
let applyM  f x = ret (apply f x)
\end{code}

@@

\begin{code}
module S(T:
  sig
	(* Representation type of values,  to be specified *)
    type ('a, 'b) rep
  end) = struct

open T
module type DOMAINL = sig
  include DOMAIN
  type 'a vc = ('a,v) rep
  val zeroL : 'a vc
  val oneL : 'a vc
  val ( +^ ) : 'a vc -> 'a vc -> 'a vc
  val ( *^ ) : 'a vc -> 'a vc -> 'a vc
  val ( -^ ) : 'a vc -> 'a vc -> 'a vc
  val uminusL : 'a vc -> 'a vc
  val divL : 'a vc -> 'a vc -> 'a vc
  val better_thanL : ('a vc -> 'a vc -> ('a,bool) rep) option
  val normalizerL : ('a vc -> 'a vc) option
end

module T = Domains_sig.S(struct type ('a, 'b) rep = ('a, 'b) code end)
open T

open Domains_common
(* because the operations are "syntactic" to a certain extent,
   we have to repeat ourselves a lot *)
module IntegerDomainL = struct
    include IntegerDomain
    type 'a vc = ('a,v) code
    let zeroL = .< 0 >.  
    let oneL = .< 1 >. 
    let (+^) x y = .<.~x + .~y>. 
    let ( *^ ) x y = .<.~x * .~y>.
    let ( -^ ) x y = .<.~x - .~y>.
    let uminusL x = .<- .~x>.
    let divL x y = .<.~x / .~y>. 
    let normalizerL = None
    let better_thanL = Some (fun x y -> .<abs .~x > abs .~y >. )
end
\end{code}

@@

\section{Runge-Kutta solvers}
\label{s:ode}

@@

There are many more aspects which can also be handled:
Input variations (augmented
matrices), error reporting (i.e. asking for the determinant of a 
non-square matrix), memory hierarchy issues, loop-unrolling
\cite{Padua:MetaOcaml:04},
warnings when zero-testing is undecidable and
a value is only probabilistically non-zero, etc.  The larger program
family of LU decompositions contains more aspects still.

@@

The actual bind function of the monad defaults to |bind| and the
match-failure function to |failwith| (only used for refutable
patterns; see below).  To select a different function, use the
extended forms of |perform|. For example, use the function named 
|bind| from module |Mod|.  In
addition use the module's |failwith|-function in refutable patterns.
\begin{code}
        perform with module Mod in exp2
\end{code}
The code has full explanation...

@@

   let t_4 = {arr = (Array.copy a_1.arr)} (a_1) in
   let t_5 = a_1.m in
   let t_6 = a_1.n in

@@

     begin (* full pivoting *)

@@

          if ((fst (fst i_16)) <> t_13) then begin (* swapping of rows?*)

@@

   (t_4,

@@

The code generated for |GenFA1|, LU of the floating point matrix

@@

val resFA1 : ('a, GAC_F.contr -> GenFA1.O.res) code =

@@

@@

    let t_10 = (! t_2) in
    let t_11 = (! t_3) in
    let t_12 = (ref (None)) in
    let t_18 =

@@

      for j_15 = t_10 to (t_7 - 1) do
       let t_16 = (t_5.(j_15)).(t_11) in
       if (t_16 <> 0.) then
        (match (! t_12) with
         | Some (i_17) ->
            if ((abs_float (snd i_17)) < (abs_float t_16)) then
             (t_12 := (Some (j_15, t_16)))

@@

         | None -> (t_12 := (Some (j_15, t_16))))

@@

      (match (! t_12) with (* swapping of rows *)
       | Some (i_13) ->
          if ((fst i_13) <> t_10) then
           let t_14 = t_5.(t_10) in
           t_5.(t_10) <- t_5.(fst i_13);
           t_5.(fst i_13) <- t_14
          else ();
          (Some (snd i_13))

@@

    (match t_18 with
     | Some (i_19) -> (* elimination loop, elided?? *)

@@

         for j_20 = (t_10 + 1) to (t_7 - 1) do
          let t_21 = (t_5.(j_20)).(t_11) in
          if (t_21 <> 0.) then begin
           for j_22 = (t_11 + 1) to (t_6 - 1) do
            (t_5.(j_20)).(j_22) <-
             ((t_5.(j_20)).(j_22) -. ((t_21 /. i_19) *. (t_5.(t_10)).(j_22)))
           (t_5.(j_20)).(t_11) <- 0.

@@

   t_5>.

@@

\begin{code2}
val resFA9 : ('a, GAC_F.contr -> GenFA9.O.res) code =
  .<fun a_1 ->
   let t_2 = (ref 0) in
   let t_3 = (ref 0) in
   let t_5 = (Array.map (fun x_4 -> (Array.copy x_4)) (Array.copy a_1)) in
   let t_6 = (Array.length a_1.(0)) in
   let t_7 = (Array.length a_1) in
   let t_8 = (ref ([])) in
   while (((! t_3) < t_6) && ((! t_2) < t_7)) do
    let t_9 = (! t_2) in
    let t_10 = (! t_3) in
    let t_11 = (ref (None)) in
    let t_17 =
     begin
      for j_14 = t_9 to (t_7 - 1) do
       let t_15 = (t_5.(j_14)).(t_10) in
       if (t_15 <> 0.) then
        (match (! t_11) with
         | Some (i_16) ->
            if ((abs_float (snd i_16)) < (abs_float t_15)) then
             (t_11 := (Some (j_14, t_15)))
            else ()
         | None -> (t_11 := (Some (j_14, t_15))))
       else ()
      done;
      (match (! t_11) with
       | Some (i_12) ->
          if ((fst i_12) <> t_9) then begin
           let t_13 = t_5.(t_9) in
           t_5.(t_9) <- t_5.(fst i_12);
           t_5.(fst i_12) <- t_13;
           (t_8 := ((RowSwap ((fst i_12), t_9)) :: (! t_8)))
          end else ();
          (Some (snd i_12))
       | None -> (None))
     end in
    (match t_17 with
     | Some (i_18) ->
        begin
         for j_19 = (t_9 + 1) to (t_7 - 1) do
          let t_20 = (t_5.(j_19)).(t_10) in
          if (t_20 <> 0.) then
           for j_21 = (t_10 + 1) to (t_6 - 1) do
            (t_5.(j_19)).(j_21) <-
             ((t_5.(j_19)).(j_21) -. ((t_20 /. i_18) *. (t_5.(t_9)).(j_21)))
           done
          else ()
         done;
         ()
        end;
        (t_2 := ((! t_2) + 1))
     | None -> ());
    (t_3 := ((! t_3) + 1))
   done;
   (t_5, (! t_8))>.
\end{code2}

>>>>

\jacques{This whole section needs to be edited to take into account
the different organization of the material}.

\subsection{Domains}
Reviewing the various aspects outlined in section~\ref{aspects}, the
simplest parametrization is to make the domain abstract.  As it turns
out, we need the following to exist in our domains: $0$, $1$, $+$,
$*$, (unary and binary) $-$, at least \emph{exact} division,
normalization, and potentially a relative size measure. The simplest
case of such domain abstraction is |param_code1|.  There,
code-generators such as |plus| and |one| were passed as arguments. We
need far more than two parameters, so we have to group them. Instead
of the grouping offered by regular records, we use OCaml
\emph{structures} (i.e., modules) so we can take advantage of
extensibility, type abstraction and constraints, and especially
parameterized structures (\emph{functors}).  We define the type of the
domain, the signature |DOMAIN|, which different domains must satisfy:

@@

module IntegerDomain : DOMAIN = struct
    type v = int
    let kind = Domain_is_Ring
    let zero = 0 and one = 1
    let plus x y = x + y  and  minus x y = x - y
    let times x y = x * y and  div x y = x / y
    let uminus x = -x
    let normalizer = None
    let better_than = Some (fun x y -> abs x > abs y)
end
\end{code}

One particular domain instance is |IntegerDomain|. The notation\\
|module IntegerDomain : DOMAIN| makes the compiler verify that our
|IntegerDomain| is indeed a |DOMAIN|, that is, satisfies the required
signature. The constraint |DOMAIN| may be omitted; in that case, the
compiler will verify the type when we try to use that structure as a
|DOMAIN|. In any case, the errors such as missing ``methods'' or
methods with incorrect types will be caught statically, even
\emph{before} any code generation takes place. The variant
|Domain_is_Ring| of |IntegerDomain.domain_kind| encodes a semantic constraint 
that the full division
is not available. While the |DOMAIN| type may have looked daunting to
some, the implementation is quite straightforward.  Other domains such
as |float| and arbitrary precision exact rational numbers |Num.num|
are equally simple.

A more complex domain is |Zp| of the field of integers in prime
modulus:
\begin{code}
module ZpMake(P:sig val p:int end) = struct
    type v = int
    let kind = Domain_is_Field
    let zero = 0 and one = 1
    let plus x y = (x + y) mod P.p
    let times x y = (x * y) mod P.p
    \dots
    let normalizer = None and better_than = None
    let () = assert (is_prime P.p)
end
\end{code}
This domain is parametrized by an integer number |p|, the
(modulus?). To be more precise, the structure |ZpMake| is
parameterized over another structure of the type described by
signature P; that argument structure has one field, the int value |p|.
Such a parameterized structure (or, a function from structures to
structures) is a \emph{functor}. The result of |ZpMake| is a domain that is a
field with no defined order. Hence
|normalizer| and |better_than| are set to |None|. |Zp| forms
a field only when |p| is prime. Therefore, we must make this check,
in the last line of the above code. That line differs from the other 
bindings in |ZpMake| in that it neither defines a function, such as
|plus|, nor binds a value, such as |zero|. Therefore, the non-value
expression |assert (is_prime P.p)| will be evaluated
when the corresponding module is instantiated, like the following
\begin{code}
module Z19 = ZpMake(struct let p = 19 end)
\end{code}
We will call such an expression an initializing expression.  if we
replace |p = 19| with |p = 9| above, we receive an error, a run-time
error. It is raised however as we instantiate and combine modules that
will eventually make the generator.  Although the error is reported at
run-time rather than compile time as one might have hoped, the error
is raised when \emph{generating the generator} -- well before the
generation of the target code could begin. In our code we make
extensive use of these so-called preflight checks performed as part of
module initialization. The checks seem to offer a good compromise:
they are dynamic and so do not require complicated type systems; on
the other hand, the checks are performed quite early, when buiding
code generators, and so ensure that no code violating the
corresponding \emph{semantic} constraints will be generated. Although
some may frown on the use of module initializing expressions (since
that requires careful attention to sharing and multiple instantiations
of a module), these concerns do not apply in our case: our preflight
tests are all idempotent and maintain no state.

\subsection{Abstract code and lifted domains}
We will be using MetaOCaml operations to build code expressions, values
of the type |('a,'v) code|. However, we wish to make our generators
abstract over the particular code representations. Therefore, in the
following Sections (in the rest of the paper?) rather than 
MetaOCaml brackets and escapes, we will be using code combinators
described below. These combinators are functions that produce and
combine values of the type of abstract code.

This abstraction lets us vary the output form of our generated LU
programs. We may use thunks rather than MetaOCaml |('a,'v) code|
values, for benchmarking purposes. We could use code combinators that
produce C or Fortran code.

Our code combinator library includes operations to create, dereference
and mutate reference cells, compare values, manipulate values used
for indexing arrays. The following is the implementation of a sample
combinators, in the case of the abstract code being MetaOCaml code values. 
\begin{code}
let lift x = .< x >.

let cunit = .< () >.
let unitL = fun s k -> k s .< () >.

let liftRef x = .< ref .~x >. 
let liftGet x = .< ! .~x >. 
let liftPair x = (.< fst .~x >., .< snd .~x >.)

module Logic = struct
  let notL a        = .< not .~a >.
  let equalL a b    = .< .~a = .~ b >.
  let notequalL a b = .< .~a <> .~ b >.
  let andL a b     = .< .~a && .~b >. 
end

module Idx = struct
  let zero = .< 0 >. and one = .< 1 >. and minusone = .< -1 >.
  let succ a = .< .~a + 1 >.
  let pred a = .< .~a - 1 >.
  let less a b = .< .~a < .~b >.
  let uminus a = .< - .~a >.
  let add a b = .< .~a + .~b >.

let update a f = let b = f (liftGet a) in .< .~a := .~b >.
let assign a b = .< .~a := .~b >.
let apply  f x = .< .~f .~x >.
let updateM a f = ret (update a f)
let assignM a b = ret (assign a b)
let applyM  f x = ret (apply f x)
\end{code}

@@

We now convert the domains of values from teh previous section into
lifted domains, of abstract code values. We define the signature
|DOMAINL| 
\begin{code}
module S(T: sig type ('a, 'b) rep  end) = struct
open T
module type DOMAINL = sig
  include DOMAIN
  type 'a vc = ('a,v) rep
  val zeroL : 'a vc
  val oneL : 'a vc
  val ( +^ ) : 'a vc -> 'a vc -> 'a vc
  val ( *^ ) : 'a vc -> 'a vc -> 'a vc
  val ( -^ ) : 'a vc -> 'a vc -> 'a vc
  val uminusL : 'a vc -> 'a vc
  val divL : 'a vc -> 'a vc -> 'a vc
  val better_thanL : ('a vc -> 'a vc -> ('a,bool) rep) option
  val normalizerL : ('a vc -> 'a vc) option
end 
\end{code}
(part of the larger module |Domains_sig|) which is parameterised by
the type of the abstract code, |('a, 'b) rep|. The line |include DOMAIN|
says that lifted domains include all members of non-lifted domains,
including the initializing expressions with pre-flight checks.
non-lifted

The following is a particular instance of |DOMAINL|, the lifted
version of |IntegerDomain| of the previous section, in the particular
case of abstract code being MetaOCaml code values |('a, 'b) code|:
\begin{code}
module T = Domains_sig.S(struct type ('a, 'b) rep = ('a, 'b) code end)
open T
module IntegerDomainL = struct
    include IntegerDomain
    type 'a vc = ('a,v) code
    let zeroL = .< 0 >.  
    let oneL = .< 1 >. 
    let (+^) x y = .<.~x + .~y>. 
    let ( *^ ) x y = .<.~x * .~y>.
    let ( -^ ) x y = .<.~x - .~y>.
    let uminusL x = .<- .~x>.
    let divL x y = .<.~x / .~y>. 
    let normalizerL = None
    let better_thanL = Some (fun x y -> .<abs .~x > abs .~y >. )
end
\end{code}

\subsection{Containers}

@@

\oleg{We should mention somewhere that we have implemented not only GE
  but Solvers as well.}

\section{Runge-Kutta solvers}
\label{s:ode}
\oleg{Drop this? no space}

@@

There are many more aspects which can also be handled:
error reporting (i.e. asking for the determinant of a 
non-square matrix), memory hierarchy issues, loop-unrolling
\cite{Padua:MetaOcaml:04},
warnings when zero-testing is undecidable and
a value is only probabilistically non-zero, etc.  The larger program
family of LU decompositions contains more aspects still.

@@

The actual bind function of the monad defaults to |bind| and the
pattern-match--failure function to |failwith| (only used for refutable
patterns).  To select a different function, use the
extended forms of |perform|. For example, use the function named 
|bind| from module |Mod|.  In
addition use the module's |failwith|-function in refutable patterns.
\begin{code}
        perform with module Mod in exp2
\end{code}
The code has full explanation...

@@

   let t_4 = (a_1) \{arr = (Array.copy a_1.arr)\} in
   let t_5 = a_1.m in  (* magnitude of det *)
   let t_6 = a_1.n in  (* sign of the det *)

@@

     begin (* full pivoting, search for the pivot *)

@@

          if ((fst (fst i_16)) <> t_13) then begin (* swapping of rows *)

@@

   (t_4, (* matrix with the U factor *)

@@

The code generated for |GenFA9|, LU of the floating point matrix

@@

val resFA9 : ('a, GAC_F.contr -> GenFA9.O.res) code =

@@

   let t_8 = (ref ([])) in  (* accumulate permutations in a list *)

@@

    let t_9 = (! t_2) in
    let t_10 = (! t_3) in
    let t_11 = (ref (None)) in
    let t_17 =

@@

      for j_14 = t_9 to (t_7 - 1) do
       let t_15 = (t_5.(j_14)).(t_10) in
       if (t_15 <> 0.) then
        (match (! t_11) with
         | Some (i_16) ->
            if ((abs_float (snd i_16)) < (abs_float t_15)) then
             (t_11 := (Some (j_14, t_15)))

@@

         | None -> (t_11 := (Some (j_14, t_15))))

@@

      (match (! t_11) with  (* swapping of rows *)
       | Some (i_12) ->
          if ((fst i_12) <> t_9) then begin
           let t_13 = t_5.(t_9) in
           t_5.(t_9) <- t_5.(fst i_12);
           t_5.(fst i_12) <- t_13;  (* and accumulate permutations *)
           (t_8 := ((RowSwap ((fst i_12), t_9)) :: (! t_8)))
          (Some (snd i_12))

@@

    (match t_17 with  (* elimination loop *)
     | Some (i_18) ->

@@

         for j_19 = (t_9 + 1) to (t_7 - 1) do
          let t_20 = (t_5.(j_19)).(t_10) in
          if (t_20 <> 0.) then
           for j_21 = (t_10 + 1) to (t_6 - 1) do
            (t_5.(j_19)).(j_21) <-
             ((t_5.(j_19)).(j_21) -. ((t_20 /. i_18) *. (t_5.(t_9)).(j_21)))

@@

   (t_5, (! t_8))>. (* return both L and U factors, list permutations *)

@@
